TRIK Studio: Technical Introduction

### 2017-02-10 2017-trikStudioEnglish/trikStudioEnglish.tex
Updated quotes
Polished constraints checking section

\verb|LuaTokenTypes::semicolon| is a token corresponding to ';', operator '-' creates a simple parser that can parse only semicolons and excludes semicolons from the AST, ``explist'' is a reference parser object, like ``stat'', defined elsewhere, '\&' combines two parsers into a parser which accepts concatenation of their corresponding strings, '|' combines two parsers into a parser that accepts alternative, '\verb|~|' creates an optional parser from its argument, which shall also be a parser. There are also operators for adding semantic actions to productions and assigning parser a name for debug purposes. When all parsers are combined in such a way, it is enough to call 'parse()' method of a resulting parser object, giving it a token stream. The parser combinator library was used for another language in QReal~\cite{tikhonova2015generation}, so it is general enough to support not only Lua.

@@

Last important subsystem which is described in this paper is automatic checker of constraints for TRIK Studio programs. This subsystem allows to turn a world model in a simulator to an exercise with specified success and failure conditions, which can be shared among students and solutions (in a form of visual programs) can be automatically checked against these success conditions in the simulator, thus providing feedback without an intervention from a teacher. To create an exercise one needs to specify two things:

@@

\begin{itemize}
    \item $id_i$ --- event id: an internal label by which other parts of a program may refer to this event, can be empty;
    \item $c_i$ --- condition, under which the event is raised, a formula in a first-order predicate calculus without quantification;
    \item $T_i$ --- ordered set of triggers $[ t_{i_1}, t_{i_2}, ..., t_{i_n} ]$. Each trigger specifies an action that shall be done when event condition $c_i$ becomes true.
\end{itemize}

Each event is specified by its XML node in a program. An event can be specified in a \textit{canonical} form, or in a form of a \textit{constraint}. An event in a canonical form is a previously described triple $(id_i, c_i, T_i)$. An event in a constraint form is a triple $(id_j, c_j, message_j)$. A constraint is interpreted as an event $(id_i, $$\neg$$c_i, [ fail(message_j) ])$, where $fail(message_j)$ is a trigger that stops simulation and reports a $message_j$ error message. In other words, a constraint in an event that is raised when given condition is violated and reports this violation to a user. Events in a form of constraints are added to a language for pragmatical reasons only, because it is much more convenient to specify conditions like ``robot $x$ shall not leave area $a$'' or ``robot $x$ shall have a set of sensors $s$ connected'' as constraints instead of events. Time limits are a special case of such constraints. A time limit shall be specified for each constraint definition program so the constraints checker will not run the simulation indefinite amount of time if all constraints are satisfied but success event is never triggered. The checker verifies that a time limit is set before execution of a program, and if not, considers it as a semantic error.

Let's briefly describe predicates,functional symbols and elementary triggers used in the language. Predicate symbols are be divided into these groups.
\begin{itemize}
    \item Comparison predicates $>$, $<$, $\leq$, $\geq$, $=$, $\neq$.
    \item Spatial predicates. They have an ``item $x$ is located inside area $y$'' form.
    \item Event state predicates $settedUp(id_i)$ and $dropped(id_i)$, denoting active and inactive events accordingly. Active events can be raised when their corresponding condition is satisfied, events in inactive state will not be raised (and their triggers will not be executed) even when condition is satisfied.
    \item Time predicate $timer(t)$, which is initially false, becomes true after $t$ ticks of model time and stays true afterwards.
    \item Other predicates that can be expressed by already described predicates, they are introduced for pragmatical reasons only.
\end{itemize}

Functional symbols are the following.
\begin{itemize}
    \item Constants of different types (integer, floating-point, string, colour, geometrical and so on).
    \item $variableValue(id)$ symbol that denotes a value of a variable with identifier $id$. Variables are internal to the constraints checking program (i.e. do not represent simulation state) and can be useful for implementing complex conditions like counting the times that robot does some action.
    \item Arithmetical and geometrical operations over other values: for example, an absolute value of a number or a distance between two points.
    \item Comparison of forms of two figures using Levenshtein distance. It is useful to compare images drawn on robot display to an ideal image specified in an exercise.
    \item $objectState(path)$ symbol allows to get state of a simulated robot devices or objects of a simulated world. $path$ argument shall contain a path to a desired property in a hierarchy of objects in world model. Such path will be translated into a sequence of C++ object property references using Qt Reflection mechanism, so this symbol provides a bridge between constraints definition and the simulator.
\end{itemize}

@@

Architecturally the constraints checking system is a standalone module which uses TRIK Studio core and the simulator model as shared libraries. The simulator model and robotic kit support in TRIK Studio are frequently extended with new devices, options and features, so Qt Reflection mechanism is used to access objects and their properties from constraint specifications. It allows to extend the TRIK Studio core without modifications in the constraints checking system itself, new properties immediately become available from the constraints description language. The constraints checker subscribes to events from the simulation model as another view (in MVC pattern sense) and checks constraints on every tick of model time. It may seem very ineffective, but only active constraints are checked and only a few constraints are active at any given time, so a simulation with constraints checking has almost the same performance as without it.

The described language proven to be quite effective at specifying spatial and temporal constraints on a system state. It is more expressive than, for example, temporal logic or topological temporal logic languages which are used for specification of formal constraints on robot behavior in recent works~(\cite{mordvinov2016formal,kress2007s,бугайченко2007разработка,дмитриев2013адаптация}). And this language is not Turing-complete (for example, it can not express Markov algorithms due to absence of means to work with arbitrary length collections, so it can not express string replacement), it opens a possibility to statically analyze non-trivial properties of programs on it. In addition to this, a subset of the language can be used for specification of requirements for a program to be used in formal verification. Plans for next versions of TRIK Studio include creating visual editor for constraints, which, combined with some verification engine, will possibly allow to formally prove correctness of visual programs without the need to actually write temporal logic formulas. Investigation of this idea is a promising future work direction.

The constraints checking system is also used for functional testing of TRIK Studio on continuous integration server. For more information on this (and other details of TRIK Studio testing) see in~\cite{mordvinov2016testing}.

This functionality is also used as an automatic checker of exercises for MOOC on Stepik platform\footnote{https://stepik.org/s/7qe3xj4Z (accessed: 07.02.2017)}. Without such a checker it would be impossible to provide feedback to possibly large number of students, making course much less interactive. From a technical point of view the checker is a set of scripts which launch ``headless'' interpreter (i.e. interpreter without GUI) on a correct field with correct constraints. Checker is hooked up into the Stepik infrastructure and runs in a separate Docker\footnote{https://www.docker.com/ (accessed: 07.02.2017)} container when a new solution is submitted. The checker has several fields and corresponding constraint descriptions for each task (from one to five), each solution is checked against all those fields to test that it works in different situations and is not hardcoded. Fields are hidden from users but error reporter output does gets displayed, so students can guess what went wrong (well, theoretically, but to our opinion it is better than traditional ACM ICPC system, where only a number of incorrect test and a general type of error are reported).

>>>>

\verb|LuaTokenTypes::semicolon| is a token corresponding to ';', operator '-' creates a simple parser that can parse only semicolons and excludes semicolons from the AST, ``explist'' is a reference parser object, like ``stat'', defined elsewhere, '\&' combines two parsers into a parser which accepts concatenation of their corresponding strings, '|' combines two parsers into a parser that accepts alternative, '\verb|~|' creates an optional parser from its argument, which shall also be a parser. There are also operators for adding semantic actions to productions and assigning parser a name for debug purposes. When all parsers are combined in such a way, it is enough to call 'parse()' method of a resulting parser object, giving it a token stream. The parser combinator library was used for anoFther language in QReal~\cite{tikhonova2015generation}, so it is general enough to support not only Lua.

@@

Last important subsystem which is described in this paper is automatic checker of constraints for TRIK Studio programs. This subsystem allows to turn a world model in a simulator to an exercise with specified success and failure conditions, which can be shared among students, and solutions (in a form of visual programs) can be automatically checked against these success conditions in the simulator, thus providing feedback without an intervention from a teacher. To create an exercise one needs to specify two things:

@@

\begin{itemize}
    \item $id_i$ --- event id: an internal label by which other parts of a program may refer to this event, can be empty;
    \item $c_i$ --- condition, under which the event is raised, a formula in a first-order predicate calculus without quantification;
    \item $T_i$ --- ordered list of triggers $[ t_{i_1}, t_{i_2}, ..., t_{i_n} ]$. Each trigger specifies an action that shall be done when event condition $c_i$ becomes true.
\end{itemize}

Each event is specified by its XML node in a program. An event can be specified in a \textit{canonical} form, or in a form of a \textit{constraint}. The event in a canonical form is a triple $(id_i, c_i, T_i)$. The event in a constraint form is a triple $(id_j, c_j, message_j)$. A constraint is interpreted as an event $(id_i, $$\neg$$c_i, [ fail(message_j) ])$, where $fail(message_j)$ is a trigger that stops simulation and reports a $message_j$ error message. In other words, a constraint is the event that is raised when given condition is violated and reports this violation to a user. Events in a form of constraints are added to a language for pragmatical reasons only, because it is much more convenient to specify conditions like ``robot $x$ shall not leave area $a$'' or ``robot $x$ shall have a set of sensors $s$ connected'' as constraints instead of events. Time limits are a special case of such constraints. A time limit shall be specified for each constraint definition program so the constraints checker will not run the simulation indefinite amount of time if all constraints are satisfied but success event is never triggered. The checker verifies that a time limit is set before execution of a program, and if not, considers it as a semantic error.

Let's briefly describe predicates, function symbols and elementary triggers used in the language. Predicate symbols are divided into these groups.
\begin{itemize}
    \item Comparison predicates $>$, $<$, $\leq$, $\geq$, $=$, $\neq$.
    \item Spatial predicates. They have an ``item $x$ is located inside area $y$'' form.
    \item Event state predicates $settedUp(id_i)$ and $dropped(id_i)$, denoting active and inactive events accordingly. Active events can be raised when their corresponding condition is satisfied, events in inactive state will not be raised (and their triggers will not be executed) even when condition is satisfied.
    \item Time predicate $timer(t)$, which is initially false, becomes true after $t$ ticks of model time and stays true afterwards.
    \item Other ``syntactic sugar''.
\end{itemize}

Function symbols are the following.
\begin{itemize}
    \item Constants of different types (integer, floating-point, string, colour, geometrical and so on).
    \item $variableValue(id)$ symbol that denotes a value of a variable with identifier $id$. Variables are internal to the constraints checking program (i.e. do not represent simulation state) and can be useful for implementing complex conditions like counting the times that robot does some action.
    \item Arithmetical and geometrical operations over other values: for example, an absolute value of a number or a distance between two points.
    \item Comparison of forms of two figures using Levenshtein distance. It is useful to compare images drawn on robot display to an ideal image specified in an exercise.
    \item $objectState(path)$ symbol allows to get state of a simulated robot devices or objects of a simulated world. $path$ argument shall contain a path to a desired property in a hierarchy of objects in world model. Such path will be translated into a sequence of C++ object property references using Qt Reflection mechanism, so this symbol provides a bridge between constraints definition and the simulator environment.
\end{itemize}

@@

Architecturally the constraints checking system is a standalone module which uses TRIK Studio core and the simulator model as shared libraries. The simulator model and robotic kit support in TRIK Studio are frequently extended with new devices, options and features, so Qt Reflection mechanism is used to access objects and their properties from constraint specifications. It allows to extend the TRIK Studio core without modifications in the constraints checking system itself, new properties immediately become available from the constraints description language. The constraints checker subscribes to events from the simulation model as another view (in MVC pattern sense) and checks constraints on every tick of model time. It may seem ineffective, but only active constraints are checked each timer tick (and only a few constraints are active at any given time), so constraints checking almost does not affect the performance.

The described language has proven to be quite effective at specifying spatial and temporal constraints on a system state. It is more expressive than, for example, temporal logic or topological temporal logic languages which are used for specification of formal constraints on robot behavior in recent works~(\cite{mordvinov2016formal,kress2007s,бугайченко2007разработка,дмитриев2013адаптация}). The constraints checking system is also used for functional testing of TRIK Studio on continuous integration server. For more information on this (and other details of TRIK Studio testing) see in~\cite{mordvinov2016testing}.

This functionality is also used as an automatic checker of exercises for MOOC on Stepik platform\footnote{https://stepik.org/s/7qe3xj4Z (accessed: 07.02.2017)}. Without such checker it would be impossible to provide feedback to possibly large number of students, making course much less interactive. From a technical point of view the checker is a set of scripts which launch ``headless'' interpreter (i.e. interpreter without GUI) on a correct field with correct constraints. Checker is hooked up into the Stepik infrastructure and runs in a separate Docker\footnote{https://www.docker.com/ (accessed: 07.02.2017)} container when a new solution is submitted. The checker has several fields and corresponding constraint descriptions for each task (from one to five), each solution is checked against all those fields to test that it works in different situations and is not hardcoded. Fields are hidden from users, but error reporter output does gets displayed, so students can guess what went wrong (in our opinion it is better than traditional ACM ICPC system, where only a number of incorrect test and a general type of error are reported).
